---
title: "**Econometrics**"
subtitle: "Ordered Logit Model in R"
author: "Sujiao (Emma) ZHAO"
date: "2021/12/10 (updated: `r Sys.Date()`)"
output: html_document
---

```{r setup, include=FALSE}
require("knitr")
knitr::opts_chunk$set(echo = TRUE)
## setting working directory
opts_knit$set(root.dir = "C:/Users/bpu058246/Desktop/Teaching/Econometria/Lectures/Data/")
```



<br />

## Reminder

- If outcome or dependent variable is binary and in the form 0/1, then use logit or probit models.

- **If outcome or dependent variable is categorical but are ordered (i.e. low to high), then use ordered logit or ordered probit models.**

- If outcome or dependent variable is categorical without any particular order, then use multinomial logit.


<br />

## Running a Ordered Logit Model

- Getting sample data

```{r include=TRUE, warning = FALSE, message=FALSE}
library(foreign)
mydata <- read.dta("https://dss.princeton.edu/training/Panel101.dta")
```


- Running a ordered logit model

```{r include=TRUE, warning = FALSE, message=FALSE}
library(MASS)
m1 <- polr(opinion ~ x1 + x2 + x3, data=mydata, Hess=TRUE)
summary(m1)
```

The argument `Hess` is required for SE.

Arrange the results in a table.

```{r include=TRUE, warning = FALSE, message=FALSE}
library(stargazer)
stargazer(m1, type="text")
```


## Odds Ratio

The Estimate column shows the coefficients in log-odds form. To get the odds ratio, you need explonentiate the logit coefficient.

```{r include=TRUE, warning = FALSE, message=FALSE}
m1.or=exp(coef(m1))
m1.or
stargazer(m1, type="text", coef=list(m1.or), p.auto=FALSE)
```

Keeping all other variables constant, when x1 increases one unit, it is 2.668 times more likely to be in a higher category. In other words, the odds of moving to a higher category in the outcome variable is 166.8% when x1 move one unit (2.668 – 1). The coefficient is significant.

## Predicted Probabilities

**Use "probs" for predicted probabilities**

```{r include=TRUE, warning = FALSE, message=FALSE}
m1.pred <- predict(m1, type="probs")
summary(m1.pred)
```
The line "Mean" shows the predicted probabilities of each category when all predictors are at their mean value.

We can also set the predictors at specific values, for example x1 and x2 at their means, and x3 = 1 and x3 = 2.

```{r include=TRUE, warning = FALSE, message=FALSE}
setup1 <- data.frame(x1=rep(mean(mydata$x1),2),
x2=rep(mean(mydata$x2),2),
x3=c(1,2))
setup1
```

```{r include=TRUE, warning = FALSE, message=FALSE}
predictm1<-predict(m1, newdata=setup1, type="probs")
setup1.probs <-cbind(setup1, predictm1)
setup1.probs
```

**Use "class" for the predicted category**

```{r include=TRUE, warning = FALSE, message=FALSE}
predictm1<-predict(m1, newdata=setup1, type="class")
setup1.class <-cbind(setup1, predictm1)
setup1.class
```


## Marginal Effects

See package [`erer`](https://cran.r-project.org/web/packages/erer/index.html)

```{r include=TRUE, warning = FALSE, message=FALSE}
# install.packages("erer")
library(erer)
x <- ocME(m1)
x
```

Type the following if you want t and p-values

```{r include=TRUE, warning = FALSE, message=FALSE}
x$out
```