---
title: "**Econometrics**"
subtitle: "R Exercises: Monte Carlo Simulation for OLS Using the Monte Carlo Package"
author: "Sujiao (Emma) ZHAO"
date: "2021/10/06 (updated: `r Sys.Date()`)"
output: html_document
---

```{r setup, include=FALSE}
# library(data.table)
# library(stargazer)
# library(dplyr)
# library(magrittr)

library(pacman)
pacman:: p_load(data.table, stargazer, dplyr, magrittr)
```


<br />

## Small Samples and Asymptotic Theory

<br />

We will implement a **Monte Carlo** simulation. The idea is to use a single equation such as

\begin{equation}
y_i=\beta_1+\beta_2 x_i + \epsilon_i
\end{equation}

for the data generation where we establish beforehand the $\beta$ parameters. For each sample of data that we create we estimate the $\beta$s. 

This is the basic idea of many **Monte Carlo** simulations where you already know the model and what you want to do is to see how well your estimators or test statistics perform given that you already know the model.


# Monte Carlo Simulations Using the MonteCarlo Package 

```{r MonteCarlo-package}

p_load(MonteCarlo)

## Define the function for our OLS estimator

ols_estimator <- function(n, beta_0, beta_1){
  
  # Generate data
  U_i = rnorm(n, mean = 0, sd = 2) # Error
  X_i = rnorm(n, mean = 5, sd = 5) # Independent variable
  Y_i = beta_0 + beta_1*X_i + U_i  # Dependent variable
  
  # Formulate data.table
  data_i = data.table(Y = Y_i, X = X_i)
  
  # Run regressions
  ols_i <- lm(Y ~ X, data = data_i)
  
  # Extract slope coefficient and save
  slope <- ols_i$coefficients[2]
  #intercept <- ols_i$coefficients[1]
  
  # Return Results in a list
  return(list("beta_1" = unname(slope)))
  #return(list("beta_0" = unname(intercept)))
  
  
}

ols_estimator(n = 100, beta_0 = 1.5, beta_1 = 2.0)

## Define parameter grid
#n_grid <- c(100, 500, 1000, 2000)   # Sample sizes
n_grid <- c(100, 500, 1000)   # Sample sizes
beta0_grid <- c(2.0)          # Intercept
beta1_grid <- c(1.5)          # Slope coefficient

## Collect parameter grids in list
param_list = list("n" = n_grid, "beta_0" = beta0_grid, "beta_1" = beta1_grid)

## Run simulation
MC_result_OLS <- MonteCarlo(func = ols_estimator,    # One iteration of estimator
                            nrep = 500,              # Number of simulations
                            param_list = param_list, # Parameter grid (Sample size and betas)
                            ncpus = 1,               # Number of computers
                            time_n_test = TRUE)      # Estimate how long simulation will take?
  


```




```{r review-results}
## View results
res <- MC_result_OLS$results
res <- res$beta_1

n_100 <- res[1,,,]   # Sample size of 100
n_500 <- res[2,,,]   # Sample size of 500
n_1000 <- res[3,,,]  # Sample size of 1000 
#n_2000 <- res[4,,,]  # Sample size of 2000 

hist(n_100, xlim = c(1.20,1.70))
hist(n_500, xlim = c(1.20,1.70))
hist(n_1000, xlim = c(1.20,1.70))
#hist(n_2000, xlim = c(1.20,1.70))

```


<br />
<br />
