---
title: "**Econometrics**"
subtitle: "R Exercises: Individual Wages"
author: "Sujiao (Emma) ZHAO"
date: "2021/09/29 (updated: `r Sys.Date()`)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("C:/Users/bpu058246/Desktop/Teaching/Econometria/Lectures/")
```


<br />

## Individual Wages

<br />


1. Read the "wages" dataset (wages1.dta) into R (using read_dta from haven). 

```{r eval=FALSE}

```

<br />

2. Inspect the data

```{r eval=FALSE}

```

<br />

3. Obtain summary statistics

```{r eval=FALSE}

```

<br />

4. Add labels to the variables

```{r eval=FALSE}
library(expss) # apply labels
wages1 = apply_labels(wages1,
  exper = "experience in years",
  male = "1 if male, 0 otherwise",
  school = "years of schooling",
  wage = "wage in 1980 dollars per hour"
)
```

<br />

5. Obtain descriptive statistics on the variable *wage*


```{r eval=FALSE}

```

<br />

6. Obtain descriptive statistics on wages by gender. 

```{r eval=FALSE}

```

<br />

7. Now, we estimate the following equation and replicate the results in table 2.1
\begin{equation}
wage_i=\beta_1+\beta_2 male_i + \epsilon_i 
\end{equation}


```{r eval=FALSE}

```

<br />

8. Next we estimate the equation in Table 2.2:
\begin{equation}
wage_i=\beta_1+\beta_2 male_i +\beta_3 school_i + \beta_4 exper_i +\epsilon_i 
\end{equation}

```{r eval=FALSE}

```


