---
title: "**Econometrics**"
subtitle: "Dummy variables"
author: "Sujiao (Emma) ZHAO"
date: "2021/10/13 (updated: `r Sys.Date()`)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("C:/Users/bpu058246/Desktop/Teaching/Econometria/Lectures/")
```


<br />


We will illustrate the use of dummy variables with an application in R. R uses factor vectors to represent dummy or categorical data. Factors can be ordered or unordered. Factor vectors are built on top of integer vectors and include a unique label for each integer. 


We start by reading a dataset.



```{r include=TRUE, warning = FALSE, message=FALSE}
setwd("C:/Users/bpu058246/Desktop/Teaching/Econometria/Lectures/Data/")
library(haven) # import Stata data
housing2 <- read_dta("housing2.dta")
```

These data contains records for over 5,000 individual house sales with information on characteristics of the house

```{r include=TRUE, warning = FALSE, message=FALSE}
library(Hmisc)
describe(housing2)
```

Next we run summary statistics


```{r include=TRUE, warning = FALSE, message=FALSE}
library(stargazer)
stargazer(as.data.frame(housing2), type = "text")
```


We start by regressing price on a constant only.



```{r include=TRUE, warning = FALSE, message=FALSE}
library(stargazer)
print("Regression 1")
reg1<-lm(price~1, data = housing2)
stargazer(reg1, type="text", digits=8)

```


A couple things are worth mentioning: the estimate for the $\beta_1$ is the mean of price and the $R^2$ is zero.

Next we will run several different specification with the single purpose of illustrating how to handle dummies in R. We start by adding a dummy variable that takes the value 1 if the house was built after 2000. First we tabulate the variable


```{r include=TRUE, warning = FALSE, message=FALSE}
#table(housing2$built01plus)
library(epiDisplay)
# he tab1() function also prints a bar chart to show relative frequencies by default.
tab1(housing2$built01plus, sort.group = "decreasing", cum.percent = TRUE)
```


and now we run the regression.


```{r include=TRUE, warning = FALSE, message=FALSE}
reg<-lm(price~built01plus, data = housing2)
# summary(reg)
# anova(reg)
library(stargazer)
stargazer(reg, type="text")
```


The estimate associated with $\beta_1$ is the average price of a house built before 2001 while the estimate for $\beta_2$ gives the difference between the average prices for the two periods.


We can easily confirm this:

```{r include=TRUE, warning = FALSE, message=FALSE}
library(dplyr)
housing2  %>%  group_by(built01plus) %>% summarise_at(vars(price), funs(mean))
```

What if I wanted to define the dummy the other way around, taking the value 1 for the period before 2001 and 0 elsewhere? 


```{r include=TRUE, warning = FALSE, message=FALSE}
housing2$before2000 <- ifelse(housing2$built01plus == 0, 1, 0)

reg<-lm(price~before2000, data = housing2)
# summary(reg)
# anova(reg)
library(stargazer)
stargazer(reg, type="text")
```



If we use a character vector as an argument in `lm`, R will treat the vector as a set of dummy variables. The number of dummies will be the number of characteristics (unique observations) minus 1.Take the variable *bedrooms*. Suppose that we want to run a regression that includes a dummy variable for each number of bedrooms. 

First, check levels in the variable *bedrooms*.
```{r include=TRUE, warning = FALSE, message=FALSE}
housing2  %>%  group_by(bedrooms) %>% summarise_at(vars(price), funs(mean))
```

Then we run regressions on the newly created factor variable *bedrooms2*.

```{r include=TRUE, warning = FALSE, message=FALSE}
housing2$bedrooms2 <- factor(housing2$bedrooms,
                       levels = c(1,2,3,4,5,6),
                       labels = c("one bedroom", "two bedrooms", "three bedrooms", "four bedrooms", "five bedrooms", "six bedrooms"))

reg<-lm(price~bedrooms2, data = housing2)
# summary(reg)
# anova(reg)
library(stargazer)
stargazer(reg, type="text", digits=4)
```



We can set a specific reference category by explicitly placing one of the levels first when specifying the level. Say now we want to use "three bedrooms" as our reference category.

```{r include=TRUE, warning = FALSE, message=FALSE}
housing2$bedrooms3 <- factor(housing2$bedrooms,
                       levels = c(3,1,2,4,5,6),
                       labels = c("three bedrooms", "one bedroom", "two bedrooms", "four bedrooms", "five bedrooms", "six bedrooms"))

reg<-lm(price~bedrooms3, data = housing2)
# summary(reg)
# anova(reg)
library(stargazer)
stargazer(reg, type="text", digits=4)
```

Another possibility is to not use a base category, but now we better drop the constant or we will have a perfect multicollinearity problem. 

```{r include=TRUE, warning = FALSE, message=FALSE}
reg<-lm(price~bedrooms2+0, data = housing2)
summary(reg)
anova(reg)
library(stargazer)
stargazer(reg, type="text", digits=4)
```



Finally, what if we want to include a dummy only "6 bedrooms"? It is very easy to create individual dummies based on a categorical variable. 


```{r include=TRUE, warning = FALSE, message=FALSE}
housing2$bedroom6<-ifelse(housing2$bedrooms == 6, 1, 0)
#housing2$bedroom6 = as.integer(housing2$bedrooms == 6)
reg<-lm(price~bedroom6+0, data = housing2)
# summary(reg)
# anova(reg)
library(stargazer)
stargazer(reg, type="text", digits=4)
```



<br />

<br />
