---
title: "**Econometrics**"
subtitle: "Difference in Differences: Panel Data"
author: "Sujiao (Emma) ZHAO"
date: "2021/12/02 (updated: `r Sys.Date()`)"
output: html_document
---

```{r setup, include=FALSE}
require("knitr")
knitr::opts_chunk$set(echo = TRUE)
## setting working directory
opts_knit$set(root.dir = "C:/Users/bpu058246/Desktop/Teaching/Econometria/Lectures/Data/")
```



<br />

##  Data Preparation

- Create a dummy variable to indicate the time when the treatment started. Lets assume that treatment started in 1994. In this case, years before 1994 will have a value of 0 and 1994+ a 1. If you already have this skip this step.

- Create a dummy variable to identify the group exposed to the treatment. In this example lets assumed that countries with code E, F, and G were treated (=1). Other countries were not treated (=0). If you already have this skip this step.

- Create an interaction between time and treated. We will call this interaction ‘did’.

```{r include=TRUE, warning = FALSE, message=FALSE}
# Getting sample data.
library(foreign)
mydata = read.dta("Panel101.dta")
mydata$time = ifelse(mydata$year >= 1994, 1, 0)
mydata$treated = ifelse(mydata$country == "E" |
  mydata$country == "F" |
  mydata$country == "G", 1, 0)
mydata$did = mydata$time * mydata$treated
```


## Estimating the DID estimator

```{r include=TRUE, warning = FALSE, message=FALSE}
library(stargazer)
didreg = lm(y ~ treated + time + did, data = mydata)
#summary(didreg)
stargazer(didreg, type="text", digits=2)
```

The coefficient for ‘did’ is the differences-in-differences estimator. The effect is significant at 10% with the treatment having a negative effect.


Now estimate the DID estimator (using the multiplication method, no need to generate the interaction)

```{r include=TRUE, warning = FALSE, message=FALSE}
didreg1 = lm(y ~ treated*time, data = mydata)
#summary(didreg1)
stargazer(didreg1, type="text", digits=2)
```

The coefficient for ‘treated#time’ is the differences-indifferences estimator (‘did’ in the previous example). The effect is significant at 10% with the treatment having a negative effect.

## Estimate the difference-in-differences effect using the fixed effects model

With panel data we can estimate the difference-in-differences effect using a fixed effects regression with unit and period fixed effects. Or equivalently we can use regression with the dependent variable in first differences:

```{r include=TRUE, warning = FALSE, message=FALSE}
library(plm)
library(lmtest)
mydata2<-pdata.frame(mydata, index=c("country", "year"))
didreg2 <- plm(y ~ treated*time, data = mydata2, index = c("country", "year"), model = "within")
didreg3<-coeftest(didreg2, vcov=function(x) vcovHC(x, cluster="group", type="HC1"))
stargazer(didreg3, type="text", digits=2)
```

<br />