---
title: "**Econometrics**"
subtitle: "R Exercises: Monte Carlo Simulation for OLS"
author: "Sujiao (Emma) ZHAO"
date: "2021/10/06 (updated: `r Sys.Date()`)"
output: html_document
---

```{r setup, include=FALSE}
# library(data.table)
# library(stargazer)
# library(dplyr)
# library(magrittr)

library(pacman)
pacman:: p_load(data.table, stargazer, dplyr, magrittr)
```


<br />

## Small Samples and Asymptotic Theory

<br />

We will implement a **Monte Carlo** simulation. The idea is to use a single equation such as

\begin{equation}
y_i=\beta_1+\beta_2 x_i + \epsilon_i
\end{equation}

for the data generation where we establish beforehand the $\beta$ parameters. For each sample of data that we create we estimate the $\beta$s. 

This is the basic idea of many **Monte Carlo** simulations where you already know the model and what you want to do is to see how well your estimators or test statistics perform given that you already know the model.


```{r OLS-n-100-M-500, include=TRUE}

## Parameters and Seed

beta_o = 1.5        # Intercept
beta_1 = 2.0        # Slope
set.seed(1)         # Seed (allows to drawn the same sequence of random numbers)
n = 100             # Sample Size
M = 500             # Number of experiments/iterations


## Storage
slope_DT <- rep(0,M)
intercept_DT <- rep(0,M)


## Begin Monte Carlo

for (i in 1:M){
  
  # Generate dat
  u_i = rnorm(n, mean = 0, sd = 2)    # Error
  X_i = rnorm(n, mean = 5, sd = 5)    # Independent Variable  
  Y_i = beta_o + beta_1*X_i + u_i     # Dependent Variable 
  
  
  # Formulate data.table
  data_i = data.table(Y = Y_i, X= X_i)
  
  # Run regressions
  ols_i <- lm(Y ~ X, data = data_i)
  
  
  # Extract slope coefficient and save
  slope_DT[i] <- ols_i$coefficients[2]
  intercept_DT[i] <-  ols_i$coefficients[1]
  
}


# Summary Statistics

estimates_DT <-data.table(beta_1 = slope_DT, beta_0 = intercept_DT)
stargazer(estimates_DT[, c("beta_1", "beta_0")], type = "text")

# visual inspection
hist(estimates_DT[, beta_1], xlim = c(1.5, 2.5))


```

Now you can play with different sample size.

<br />
<br />
