---
title: "**Econometrics**"
subtitle: "Autocorrelation: Quarter data car"
author: "Sujiao (Emma) ZHAO"
date: "2021/10/27 (updated: `r Sys.Date()`)"
output: html_document
---

```{r setup, include=FALSE}
require("knitr")
knitr::opts_chunk$set(echo = TRUE)
## setting working directory
opts_knit$set(root.dir = "C:/Users/bpu058246/Desktop/Teaching/Econometria/Lectures/Data/")
```



<br />

First, we need to change to the folder where the dataset is located.

<br />

We need to read the auto_quarter_data dataset. $gas$ is the natural log of real per capita expenditure on gasoline. $price$ is the natural log of real price on gasoline. $income$ is the natural log of real per capita disposable personal income. $miles$ is the natural log of miles per gallon.

<br />

```{r include=TRUE, warning = FALSE, message=FALSE}
library(haven) # import Stata data
auto <- read_dta("auto_quarter_data.dta")
```

<br />

We can do some descriptive statistics and show the first 10 observations:

```{r include=TRUE, warning = FALSE, message=FALSE}
library(stargazer)
stargazer(as.data.frame(auto), type = "text")
```

<br />

```{r include=TRUE, warning = FALSE, message=FALSE}
head(auto, 10)
```

<br />

We need to transform the data in “time series”:

```{r include=TRUE, warning = FALSE, message=FALSE}
library(dplyr)
auto<-auto %>% mutate(t = dense_rank(quarter))
auto
```

<br />

Suppose we want to estimate the following model

$$gas_t=β_0+β_1 income_t+β_2 price_t+u_t$$

and wish to test whether the disturbances are autocorrelated.


```{r include=TRUE, warning = FALSE, message=FALSE}
library(stargazer)
reg1<-lm(gas~ income +price, data=auto)
stargazer(reg1, type="text", digits=8)
```

Keep residuals and constructs the lagged value of residuals.


```{r include=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
auto$res1<-reg1$residuals
auto<-auto %>% mutate(res1lag1 = lag(res1))
auto
```


```{r include=TRUE, message=FALSE, warning=FALSE}
library(ggplot2)
fig1<-ggplot(data = auto, aes(y = res1, x = t)) + geom_point(col = 'blue') + geom_abline(slope = 0)
fig2<-ggplot(data = auto, aes(y = res1, x = res1lag1)) + geom_point(col = 'blue') + geom_abline(slope = 0)
library(grid)
# Create a new page
grid.newpage()
# Next push the vissible area with a layout of 1 row and 2 columns using pushViewport()
pushViewport(viewport(layout = grid.layout(1,2)))

print(fig1, vp = viewport(layout.pos.row = 1, layout.pos.col = 1))
print(fig2, vp = viewport(layout.pos.row = 1, layout.pos.col = 2))

```

The `acf()` function in the R stats package computes and plots the autocorrelation functions of a time series.The broken horizontal lines on the plots correspond to approximate 95% confidence limits. 
```{r include=TRUE, message=FALSE, warning=FALSE}
library(stats)
acf(residuals(reg1))
```

The graphs indicate positive first order autocorrelation of residuals.



## Checking autocorrelation

Now let us look at a couple of commands that test for autocorrelation.

```{r include=TRUE, message=FALSE, warning=FALSE}
library(lmtest)
dwtest(reg1)
```

```{r include=TRUE, message=FALSE, warning=FALSE}
library(lmtest)
bgtest(reg1)
```

So, in this case, there is evidence of autocorrelation.

## Dealing with autocorrelation

<br />

$\mbox{1. Estimate by OLS and use the Newey-West estimator}$

```{r include=TRUE, message=FALSE, warning=FALSE}
library(sandwich)
# Normal SE
coeftest(reg1)
# Newey West SE
coeftest(reg1, vcov = NeweyWest)
```

<br />


$\mbox{2. Use Feasible Weighted Least Squares}$


```{r include=TRUE, message=FALSE, warning=FALSE}
library(nlme)
mod.gls <- gls(gas~ income + price,
data=auto, correlation=corAR1(form = ~ t))
stargazer(mod.gls, type="text", digits=8)
```


<br />

<br />
