---
title: "**Econometrics**"
subtitle: "3.6 Illustration: Explaining Individual Wages"
author: "Sujiao (Emma) ZHAO"
date: "2021/10/13 (updated: `r Sys.Date()`)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("C:/Users/bpu058246/Desktop/Teaching/Econometria/Lectures/")
```


<br />


We start by reading the "bwages" dataset into Stata.

```{r include=TRUE, warning = FALSE, message=FALSE}
setwd("C:/Users/bpu058246/Desktop/Teaching/Econometria/Lectures/Data/")
library(haven) # import Stata data
wages1 <- read_dta("bwages.dta")
```

Next we add labels to the variables


```{r include=TRUE, warning = FALSE, message=FALSE}
library(expss) # apply labels
wages1 = apply_labels(wages1,
                      wage = "before-tax hourly wage rate, in euros per hour",
                      male = "1 if male, 0 otherwise",
                      educ = "education level",
                      exper = "experience in years"
)

wages1$male2 <- factor(wages1$male,
                       levels = c(1,0),
                       labels = c("male", "female"))

wages1$educ2 <- factor(wages1$educ,
                       levels = c(1,2,3,4,5),
                       labels = c("primary school", "lower vocational training", "intermediate level", "higher vocational training", "university level"))
```



```{r include=TRUE, warning = FALSE, message=FALSE}
library(Hmisc)
describe(wages1)
```

And now inspect the data.

```{r include=TRUE, warning = FALSE, message=FALSE}
head(wages1, 10)
```

Next we obtain the results similar to table 3.6

```{r include=TRUE, warning = FALSE, message=FALSE}
library(dplyr)
library(stargazer)
library(doBy)

wstat<-as.data.frame(wages1[, c(1,3,4,7,8)])

summaryBy(wage + educ + exper ~ male2, data = wstat,
          FUN = function(x) { c(mean = mean(x), std = sd(x)) } )



# Alternative solutions

# library(purrr)
# wstat %>% split(. $male2) %>% walk(~ stargazer(., type = "text"))

# wstat %>% split(. $male2) %>% map(summary)

# by(wstat, wstat$male2, summary)
# by(wstat[,1:3], wstat$male2, stargazer, type = 'text')
# 
# stargazer(subset(wstat, wstat$male2=="male"), title="Male", type = "text")
# stargazer(subset(wstat[,1:3], wstat$male2=="female"), title="Female", type = "text")
# 


```

and to obtain Table 3.7 we do


```{r include=TRUE, warning = FALSE, message=FALSE}
reg<-lm(wage~male + educ + exper, data = wages1)
summary(reg)
anova(reg)
library(stargazer)
stargazer(reg, type="text")
```

In table 3.8 we add $exper^2$ as an additional regressor.



```{r include=TRUE, warning = FALSE, message=FALSE}
reg<-lm(wage~male + educ + exper + I(exper^2), data = wages1)
summary(reg)
anova(reg)
library(stargazer)
stargazer(reg, type="text")
```


Now plot the residual versus fitted plot as in Figure 3.1


```{r include=TRUE, warning = FALSE, message=FALSE}
library(broom)
library(ggplot2)

augmented_reg <- augment(reg)

ggplot(augmented_reg, aes(x = .fitted, y = .resid), main="Residual Versus Fitted Value") + 
  geom_point()
```



We now proceed to estimation of the model in Table 3.9. The dependent variable is now the logarithm of wage.



```{r include=TRUE, warning = FALSE, message=FALSE}
reg<-lm(lnwage~male + lneduc + lnexper + I(lnexper^2), data = wages1)
summary(reg)
anova(reg)
library(stargazer)
stargazer(reg, type="text", digits=4)
```


and we replicate Figure 3.2


```{r include=TRUE, warning = FALSE, message=FALSE}
library(broom)
library(ggplot2)

augmented_reg <- augment(reg)

ggplot(augmented_reg, aes(x = .fitted, y = .resid), main="Residual Versus Fitted Value") + 
  geom_point()
```



And next we implement the test on (3.41)

```{r include=TRUE, warning = FALSE, message=FALSE}
library(car)
linearHypothesis(reg,c("lnexper = 0","I(lnexper^2) = 0"), test="F")
```


To replicate Table 3.11 we do

```{r include=TRUE, warning = FALSE, message=FALSE}
is.factor(wages1$educ2)
reg<-lm(lnwage~male + educ2 + lnexper, data = wages1)
summary(reg)
anova(reg)
library(stargazer)
stargazer(reg, type="text")
```

You can also use the factor function within the lm function, saving the step of creating the factor variable first.

```{r include=TRUE, warning = FALSE, message=FALSE}
reg<-lm(lnwage~male + factor(educ) + lnexper, data = wages1)
summary(reg)
anova(reg)
library(stargazer)
stargazer(reg, type="text", digits=4)
```


Next we estimate the regression in Table 3.12

```{r include=TRUE, warning = FALSE, message=FALSE}
reg<-lm(lnwage~ male2 * educ2 + male2 * lnexper, data = wages1)
# reg<-lm(lnwage~ factor(male) * factor(educ) + factor(male) * lnexper, data = wages1)
summary(reg)
anova(reg)
library(stargazer)
stargazer(reg, type="text")
```


And we test if all interactions with male are identical to 0.


```{r include=TRUE, warning = FALSE, message=FALSE}
library(car)
## test all interaction coefficients equal to 0
coefs <- names(coef(reg))
#coefs
linearHypothesis(reg,coefs[grep(":", coefs)], verbose=TRUE, test="F")
```


<br />

<br />
